---
title: "Noma Metagenomics: Short-read taxonomic based metagenomic analysis of noma swab vs saliva samples in R"
author: "Richard Goodman"
date: "`r Sys.Date()`"
output: html_document
---
This analysis was used in the short-read  (Illumina) taxonomic based metagenomic analysis of swab and saliva samples taken from the same patients with noma. The analysis was conducted in R and the paper is posted as a preprint on bioRxiv:

**Shotgun metagenomic analysis of the oral microbiomes of children with noma reveals a novel disease-associated organism**

Michael Olaleye, Angus M O'Ferrall, Richard N. Goodman, Deogracia W Kabila, Miriam Peters, Gregoire Falq, Joseph Samuel, Donal Doyle, Diana Gomez, Gbemisola Oloruntuyi, Shafiu Isah, Adeniyi S Adetunji, Elise N. Farley, Nicholas J Evans, Mark Sherlock, Adam P. Roberts, Mohana Amirtharajah, Stuart Ainsworth

bioRxiv 2025.06.24.661267; doi: https://doi.org/10.1101/2025.06.24.661267

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Getting Started in R 

## 1.1  Installing and Loading Packages

Install all necessary packages into R.

```{r, message=FALSE, warning=FALSE}

# Make sure these are all installed as packages first 
# Load necessary libraries
library(readxl)
library(pheatmap)
library(dplyr)
library(tidyr)
library(tibble)
library(RColorBrewer)
library(miaViz)
library(scater)
library(mia)
library(TreeSummarizedExperiment)
library(here)
library(readr)
library(phyloseq)
library(DESeq2)
library(ggrepel)
library(gridExtra)
library(vegan)
```

# 2. Import and Clean Data

We will be importing MetaPhlan style Bracken data 

First we'll write a function to import MetaPhlan style bracken data 

## 2.2 Load taxonomic data

```{r, message=FALSE}

file_path =  "../data/noma_HMP_saliva_bracken_MetaPhlan_style_report_bacteria_only_A1_A40_plus_A14.txt"

# Import data
tse_metaphlan_sal_swb = loadFromMetaphlan(file_path)

# Defining the TSE for the rest of the script
tse_metaphlan_sal_swb

```


## 2.2 Add metadata

```{r}

patient_metadata = read_excel("../data/micro_study_metadata.xlsx")
sample_to_patient = read_excel("../data/sample_to_patient_A1_A40.xlsx")

metadata = dplyr::inner_join(patient_metadata, sample_to_patient, by = "respondent_id")

metadata_2 = metadata %>% filter(sample_name  %in% colnames(tse_metaphlan_sal_swb))

View(metadata_2)

coldata = data.frame(sample_name = colnames(tse_metaphlan_sal_swb))

metadata_3 = dplyr::left_join(coldata, metadata_2, by = "sample_name")

# Create a DataFrame with this information
metadata_df = DataFrame(metadata_3)
rownames(metadata_df) = metadata_3$sample_name
t_metadata_df = t(metadata_df)
ncol(t_metadata_df)
colData(tse_metaphlan_sal_swb)
colnames(metadata_df)

# Coun specific metadata

# count saliva and swab samples
metadata_3 %>% group_by(sample_type) %>% summarise(Count = n())
# A tibble: 2 × 2
#sample_type Count
#<chr>       <int>
#  1 saliva         17
#  2 swab           11

# count participants with both swab and saliva
metadata_3 %>% group_by(respondent_id) %>% summarise(Count = n())
metadata_3 %>% group_by(respondent_id) %>% summarise(Count = n()) %>% filter(Count == 2)
metadata_3 %>% group_by(respondent_id) %>% summarise(Count = n()) %>% filter(Count == 2) %>% summarise(Count = n())

# next  pull out respondants with both saliva and swab data
metadata_3 %>% group_by(respondent_id) %>% summarise(Count = n()) %>% filter(Count == 2)
swb_and_sal = metadata_3 %>% group_by(respondent_id) %>% summarise(Count = n()) %>% filter(Count == 2)
ids_to_remove = swb_and_sal %>% pull(respondent_id)
swb_and_sal_filtered = metadata_3  %>% filter(!respondent_id %in% ids_to_remove)

# count saliva and swab samples without particpants above
swb_and_sal_filtered %>% group_by(sample_type) %>% summarise(Count = n())

# Print results
print(ids_to_remove)   # Output as a character vector

# Add this DataFrame as colData to your TreeSummarizedExperiment object
colData(tse_metaphlan_sal_swb) = metadata_df
```


## 2.3 Inspecting the Data

```{r, echo=FALSE, eval=FALSE}

(count = assays(tse_metaphlan_sal_swb)[[1]])

rowData(tse_metaphlan_sal_swb)

colData(tse_metaphlan_sal_swb)

metadata(tse_metaphlan_sal_swb)

```

## 2.4 Converting TSE to other common data formats e.g. Phyloseq 

```{r}
# Use makePhyloseqFromTreeSE from Miaverse

phyloseq_metaphlan = makePhyloseqFromTreeSE(tse_metaphlan_sal_swb)
```



# 3. Non-parametric statistical tests 

## 3.1 Preparing the data

```{r}
# See above "Converting TSE to other common data formats e.g. Phyloseq"
# Use makePhyloseqFromTreeSE from Miaverse

# make an assay for abundance
tse_metaphlan_sal_swb = transformAssay(tse_metaphlan_sal_swb, assay.type="counts", method="relabundance")

taxonomyRanks(tse_metaphlan_sal_swb)

# make an altExp and matrix for order
altExp(tse_metaphlan_sal_swb,"Genus")

tse_metaphlan_sal_swb_genus = altExp(tse_metaphlan_sal_swb, "Genus")

# Check that colData was added successfully
colData(tse_metaphlan_sal_swb_genus) = metadata_df

metadata_noma_genus = as.data.frame(colData(tse_metaphlan_sal_swb_genus))

# species
phyloseq_noma = makePhyloseqFromTreeSE(tse_metaphlan_sal_swb)
# genus
phyloseq_noma = makePhyloseqFromTreeSE(tse_metaphlan_sal_swb_genus)

phyloseq_noma_esd = transform_sample_counts(phyloseq_noma, function(x) 1E6 * x/sum(x))

ntaxa(phyloseq_noma_esd) 
nsamples(phyloseq_noma_esd) 
```


## 3.2 Permanova across entire dataset for age, sex, sample type and noma stage

```{r}

set.seed(123456)

# Calculate bray curtis distance matrix on main variables 
noma.bray = phyloseq::distance(phyloseq_noma_esd, method = "bray")
sample.noma.df = data.frame(sample_data(phyloseq_noma_esd))
permanova_all = vegan::adonis2(noma.bray ~ sex , data = sample.noma.df)
permanova_all
```

Next we will test the beta dispersion 

```{r}
# All together now
vegan::adonis2(noma.bray ~ sample_type, data = sample.noma.df)
beta = betadisper(noma.bray, sample.noma.df$sample_type)
permutest(beta)
# we don't want this to be significant 
```


## 3.3 Anosim across entire dataset

```{r}
condition_group = get_variable(phyloseq_noma_esd, "respondent_id")
set.seed (123456)
anosim(distance(phyloseq_noma_esd, "bray"), condition_group)
condition_ano = anosim(distance(phyloseq_noma_esd, "bray"), condition_group)
condition_ano
```


## 3.4 Running non-parametric tests across several variables 


```{r}

colnames((colData(tse_metaphlan_sal_swb_genus)))

# Define the list of metadata variables you want to test
variables_to_test = c("sex", "age", "noma_stage_on_admission", "respondent_id", "sample_type")

# Set a seed for reproducibility of permutation-based tests
set.seed(123456)

bray_dist = phyloseq::distance(phyloseq_noma_esd, method = "bray")

# Extract the sample data into a data frame for use with adonis2
sample_df = data.frame(sample_data(phyloseq_noma_esd))

# Create an empty list to store the results from each iteration
results_list = list()

# Loop through each variable name in the 'variables_to_test' vector
for (variable in variables_to_test) {
  
  message(paste("Running tests for variable:", variable))
  
  # PERMANOVA (adonis2) 
  # Create the statistical formula dynamically for the current variable
  formula = as.formula(paste("bray_dist ~", variable))
  
  # Run the PERMANOVA test using the adonis2 function
  permanova_res = vegan::adonis2(formula, data = sample_df, permutations = 999)
  
  # Extract the p-value from the results. It's in the 'Pr(>F)' column.
  p_permanova = permanova_res$`Pr(>F)`[1]
  
  # ANOSIM 
  # Get the grouping factor (the actual variable data) from the phyloseq object
  grouping_factor = phyloseq::get_variable(phyloseq_noma_esd, variable)
  
  # Run the ANOSIM test
  anosim_res = vegan::anosim(bray_dist, grouping_factor, permutations = 999)
  
  # Extract the p-value (significance) from the ANOSIM result
  p_anosim = anosim_res$signif
  
  # MRPP 
  # The grouping factor is the same as for ANOSIM
  # Run the MRPP test
  mrpp_res = vegan::mrpp(bray_dist, grouping_factor, permutations = 999)
  
  # Extract the p-value from the MRPP result
  p_mrpp = mrpp_res$Pvalue
  
  # Store Results 
  # Store the p-values for the current variable in our results list.
  # We create a small data frame for this iteration's results.
  results_list[[variable]] = data.frame(
    Variable = paste0(variable, "."),
    `permanova.` = p_permanova,
    `anosim.` = p_anosim,
    `mrpp.` = p_mrpp,
    # 'check.names = FALSE' prevents R from changing our column names
    check.names = FALSE 
  )
}

# Combine the list of individual data frames into one final table
final_results_table = do.call(rbind, results_list)

# Clean up the row names of the final table
rownames(final_results_table) = NULL

# Print the final, consolidated table to the console
print(final_results_table)

write.csv(final_results_table, file = "../tbls/Table_1B.csv")
```


### 3.5 Wilcox test function with FDR correction for specific taxa

```{r}

# Extract the counts and taxonomic table
counts = assay(tse_metaphlan_sal_swb_genus, "counts")
tax_table = rowData(tse_metaphlan_sal_swb_genus)$Genus  # Replace "Genus" with your taxonomic level of interest
sample_data = colData(tse_metaphlan_sal_swb_genus)

groups = as.data.frame(sample_data)

# Aggregate counts by Genus
aggregated_counts = rowsum(counts, tax_table)

# Create a new aggregated TreeSummarizedExperiment object
tse_aggregated = TreeSummarizedExperiment(assays = list(counts = aggregated_counts), 
                                           colData = sample_data)

# Calculate relative abundances
relative_abundances = sweep(assay(tse_aggregated, "counts"), 2, colSums(assay(tse_aggregated, "counts")), FUN = "/") * 100

# Convert to a data frame and group by Treatment
relative_df = as.data.frame(t(relative_abundances))

# Combine abundance data with metadata
# We add the sample names as a column to join with the metadata
full_data = relative_df %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample_id") %>%
  # Assuming your 'groups' df has rownames that are the sample_ids
  dplyr::left_join(groups %>% rownames_to_column(var = "sample_id"), by = "sample_id")

# Reshape the data from long to wide format for paired testing
# Each row will now represent one participant
wide_data = full_data %>%
  select(respondent_id, sample_type, starts_with("g__")) %>%
  pivot_wider(
    names_from = sample_type,
    values_from = starts_with("g__")
  )

# Loop through each genus, perform a paired Wilcoxon test, and store the p-value
# Define the vector of genera names (with the "g__" prefix)
genera_prefixed = paste0("g__", c("Prevotella", "Treponema", "Neisseria", "Bacteroides",
                                  "Filifactor", "Porphyromonas", "Fusobacterium", "Escherichia",
                                  "Selenomonas", "Aggregatibacter", "Capnocytophaga"))

# Initialize an empty list to store results
wilcox_results = list()

for (genus in genera_prefixed) {
  # Define the column names for the swab and saliva samples for the current genus
  saliva_col = paste0(genus, "_saliva") 
  swab_col = paste0(genus, "_swab")     
  
  # Check if both columns exist in the wide_data frame
  if (saliva_col %in% names(wide_data) && swab_col %in% names(wide_data)) {
    # Perform the paired Wilcoxon signed-rank test
    test_result = wilcox.test(
      wide_data[[saliva_col]],
      wide_data[[swab_col]],
      paired = TRUE
    )
    
    # Store the p-value
    wilcox_results[[genus]] = test_result$p.value
  }
}

# Convert the list of p-values to a data frame and apply FDR correction
fdr_wilcox_results_df = data.frame(
  Genus = gsub("g__", "", names(wilcox_results)),
  raw_p_value = unlist(wilcox_results)
)

fdr_wilcox_results_df = fdr_wilcox_results_df %>%
  mutate(
    # Apply Benjamini-Hochberg (FDR) correction
    q_value_BH = p.adjust(raw_p_value, method = "BH")
  )

# Print the final results table
print(fdr_wilcox_results_df)


print(min(fdr_wilcox_results_df$q_value_BH))
print(max(fdr_wilcox_results_df$q_value_BH))

```



# 4. Relative Abundance 

The top 20 most abundant genera were selected from across the entire dataset and visualised with the plotAbundance function of miaViz.

## 4.1 Plotting relative abundance of genera across samples 

```{r, warning=FALSE}
  
 # Check taxonomy ranks 
  taxonomyRanks(tse_metaphlan_sal_swb)
  
  # make an assay for abundance
  tse_metaphlan_sal_swb = transformAssay(tse_metaphlan_sal_swb, assay.type="counts", method="relabundance")
  
  # make an altExp and matrix for Genus
  altExp(tse_metaphlan_sal_swb,"Genus") = agglomerateByRank(tse_metaphlan_sal_swb,"Genus")
  
  # make a dataframe of relative abundance 
  relabundance_df_Genus = as.data.frame(assay(altExp(tse_metaphlan_sal_swb, "Genus"), "relabundance"))
  
  # make a matric of relative abundance 
  relabundance_matrix_Genus = assay(altExp(tse_metaphlan_sal_swb, "Genus"), "relabundance")
  
  # calculate the total relative abundance of each Genus (row sums)
  total_relabundance_Genus = rowSums(relabundance_matrix_Genus)
  
  # Identify the top 20 top Genuss
  top_Genus = names(sort(total_relabundance_Genus, decreasing = TRUE)[1:20])
  # Delete everything from start to Genus
  top_Genus = sub(".*_g__","",top_Genus)
  # Add Genus back in 
  top_Genus = paste0(paste(rep("g__", length(top_Genus)), top_Genus))
  # Delete the space introduced by this 
  top_Genus = sub(" ","",top_Genus)
  top_Genus
  
  # make a new tse_metaphlan_sal_swb where the top 14 Genuss are recognised, while others are "other"
  tse_metaphlan_sal_swb_top_20_Genus = tse_metaphlan_sal_swb
  
  head(rowData(tse_metaphlan_sal_swb_top_20_Genus)$Genus)
  
  rowData(tse_metaphlan_sal_swb_top_20_Genus)$Genus = ifelse(rowData(tse_metaphlan_sal_swb_top_20_Genus)$Genus %in% top_Genus, rowData(tse_metaphlan_sal_swb_top_20_Genus)$Genus, "-other")
  
  genus_colors = c(
    "-other" = "#E41A1C",
    "g__Actinomyces" = "#377EB8",
    "g__Aggregatibacter" = "#4DAF4A",
    "g__Bacteroides" = "#984EA3",
    "g__Campylobacter" = "#FF7F00",
    "g__Capnocytophaga" = "#FFFF33",
    "g__Dialister" = "#E7298A",
    "g__Escherichia" = "#A65628",
    "g__Filifactor" = "#F781BF",
    "g__Fusobacterium" = "#999999",
    "g_Gemella" = "#1B9E77",
    "g__Haemophilus" = "#D95F02",
    "g__Leptotrichia" = "#7570B3",
    "g__Neisseria" = "#E7298A",
    "g__Porphyromonas" = "#66A61E",
    "g__Prevotella" = "#E6AB02",
    "g__Rothia" = "#66C2A5",
    "g__Schaalia" = "#FC8D62",
    "g__Selenomonas" = "#8DA0CB",
    "g__Streptococcus" = "#E78AC3",
    "g__Tannerella" = "#E41A1C",
    "g__Treponema" = "#A6D854",
    "g__Veillonella" = "#FFD92F"
  )
  
  Genus_plot_sal_swb = plotAbundance(tse_metaphlan_sal_swb_top_20_Genus, 
                              assay.type = "relabundance", 
                              rank = "Genus", 
                              add_x_text = TRUE) +
    theme(plot.margin = ggplot2::margin(t = 30, r = 10, b = 10, l = 10))  
  

   Genus_plot_sal_swb_cols =  Genus_plot_sal_swb + scale_fill_manual(values=genus_colors)
   Genus_plot_sal_swb_cols
   
  # Order by ID 
   
  metadata_3$sample_name
  
  metadata_ID_order = metadata_3 %>% arrange(respondent_id)
  
  metadata_ID_order[,1:2]
  
  metadata_ID_order$sample_name
  
  Ordered_Genus_plot = Genus_plot_sal_swb_cols + scale_x_discrete(limits = metadata_ID_order$sample_name)

  ggsave("../imgs/Figure_1A.png", plot =  Ordered_Genus_plot, width = 28, height = 16, dpi = 400)
  
  
```

# 5. Differential Analysis with Deseq

Differential analysis used the DESeq2 model on normalised count data and determined fold-change and significant differences between the variables of the noma samples, such as noma stage, age and sex at the genera level. 

We concentrated on noma stage for this analysis, there are four stages to noma 

Stage 1: Gingivitis
Stage 2: Oedema
Stage 3: Gangrenous
Stage 4: Scarring stage 

## 5.1 Preparing the data

```{r, message=FALSE, warning=FALSE}
tse_metaphlan_sal_swb

metadata_noma = as.data.frame(colData(tse_metaphlan_sal_swb))

unique(metadata_noma$noma_stage_on_admission)


#__________________________Makes into Genus________________________________________________

# make an assay for abundance
tse_metaphlan_sal_swb = transformAssay(tse_metaphlan_sal_swb, assay.type="counts", method="relabundance")

taxonomyRanks(tse_metaphlan_sal_swb)

# make an altExp and matrix for order
altExp(tse_metaphlan_sal_swb,"Genus")

tse_metaphlan_sal_swb_genus = altExp(tse_metaphlan_sal_swb, "Genus")

# Check that colData was added successfully
colData(tse_metaphlan_sal_swb_genus) = metadata_df

metadata_noma_genus = as.data.frame(colData(tse_metaphlan_sal_swb_genus))

# Genus level
# Use makePhyloseqFromTreeSE from Miaverse

phyloseq_metaphlan_noma = makePhyloseqFromTreeSE(tse_metaphlan_sal_swb_genus)

deseq2_metaphlan_noma = phyloseq::phyloseq_to_deseq2(phyloseq_metaphlan_noma, design = ~noma_stage_on_admission)

#__________Remove or edit if other taxonomy class is needed_________________________________________

# Species level
# Use makePhyloseqFromTreeSE from Miaverse

phyloseq_metaphlan_noma = makePhyloseqFromTreeSE(tse_metaphlan_sal_swb)

deseq2_metaphlan_noma = phyloseq::phyloseq_to_deseq2(phyloseq_metaphlan_noma, design = ~noma_stage_on_admission)

# Genus level
# Use makePhyloseqFromTreeSE from Miaverse

phyloseq_metaphlan_noma = makePhyloseqFromTreeSE(tse_metaphlan_sal_swb_genus)

deseq2_metaphlan_noma = phyloseq::phyloseq_to_deseq2(phyloseq_metaphlan_noma, design = ~noma_stage_on_admission)
```


## 5.2 Differential analysis with Deseq

```{r, message=FALSE, warning=FALSE}

dds_noma = deseq2_metaphlan_noma

design(dds_noma) = ~ noma_stage_on_admission  # Replace with your column name for condition

# Run DESeq2 analysis
dds_stage = DESeq(dds_noma)
```


## 5.3  Extract results for noma stage

```{r}

unique(metadata_noma$noma_stage_on_admission)


# Extract results for diseased vs healthy
res_stage = results(dds_stage)
res_stage

string = "NA_p__Actinobacteria_c__Actinobacteria_o__Streptomycetales_f__Streptomycetaceae_g__Streptomyces"
string_result = gsub(".*(g__Streptomyces)", "\\1", string)
print(string_result)

# Clean up genus names for dds
rownames(dds_stage) = gsub(".*(g__*)", "\\1", rownames(dds_stage))

# Clean up genus names for res
res_stage@rownames = gsub(".*(g__*)", "\\1", res_stage@rownames)


res_stage_1_2 = results(dds_stage, contrast = c("noma_stage_on_admission", "Stage_1", "Stage_2"))
res_stage_1_2
res_stage_1_2[order(res_stage_1_2$padj),]

res_stage_2_3 = results(dds_stage, contrast = c("noma_stage_on_admission", "Stage_2", "Stage_3"))
res_stage_2_3

res_stage_3_4 = results(dds_stage, contrast = c("noma_stage_on_admission", "Stage_3", "Stage_4"))
res_stage_3_4

head(results(dds_stage, tidy=TRUE))
summary(res_stage)

res_stage_ordered = res_stage[order(res_stage$padj),]
head(res_stage_ordered, n =20)

res_stage_ordered_df = as.data.frame(res_stage_ordered)

res_stage_1_2_ordered = res_stage[order(res_stage_1_2$padj),]
head(res_stage_1_2_ordered, n =20)


res_stage_2_3_ordered = res_stage[order(res_stage_2_3$padj),]
head(res_stage_2_3_ordered, n =20)

res_stage_3_4_ordered = res_stage[order(res_stage_3_4$padj),]
head(res_stage_2_3_ordered, n =20)

summary(res_stage)

```

## 5.4 Inspect genera that are significantly different between noma stages

```{r}

significant_stage = as.data.frame(res_stage) %>%
  filter(padj < 0.05)
head(significant_stage) 


 significant_stage_1_2 = as.data.frame(res_stage_1_2) %>%
  filter(padj < 0.05)
head(significant_stage_1_2) 


significant_stage_2_3 = as.data.frame(res_stage_2_3) %>%
  filter(padj < 0.05)
head(significant_stage_2_3) 

significant_stage_3_4 = as.data.frame(res_stage_3_4) %>%
  filter(padj < 0.05)
head(significant_stage_3_4) 

# Order the results 

sig_res_stage = significant_stage[order(significant_stage$padj),]
sig_res_stage$genus = rownames(sig_res_stage)
head(sig_res_stage)
head(sig_res_stage, n= 15)

# Order the results for stages 1 to 2 
sig_res_stage_1_2 = significant_stage_2_3[order(significant_stage_1_2$padj),]
sig_res_stage_1_2$genus = rownames(sig_res_stage_1_2)
head(sig_res_stage_1_2)
head(sig_res_stage_1_2, n= 15)
nrow(sig_res_stage_1_2)

# Order the results for stages 2 to 3
sig_res_stage_2_3 = significant_stage_2_3[order(significant_stage_2_3$padj),]
sig_res_stage_2_3$genus = rownames(sig_res_stage_2_3)
head(sig_res_stage_2_3)
head(sig_res_stage_2_3, n= 15)
nrow(sig_res_stage_2_3)

# Order the results for stages 3 to 4 
sig_res_stage_3_4 = significant_stage_2_3[order(significant_stage_3_4$padj),]
sig_res_stage_3_4$genus = rownames(sig_res_stage_3_4)
head(sig_res_stage_3_4)
head(sig_res_stage_3_4, n= 15)
nrow(sig_res_stage_3_4)

```


## 5.5 Look for highly abundant significant ones

```{r}

taxonomyRanks(tse_metaphlan_sal_swb)

# make an altExp and matrix for Genus
altExp(tse_metaphlan_sal_swb,"Genus") = agglomerateByRank(tse_metaphlan_sal_swb,"Genus")

# make a dataframe of relative abundance 
relabundance_df_Genus = as.data.frame(assay(altExp(tse_metaphlan_sal_swb, "Genus"), "relabundance"))

# make a matric of relative abundance 
relabundance_matrix_Genus = assay(altExp(tse_metaphlan_sal_swb, "Genus"), "relabundance")

# calculate the total relative abundance of each Genus (row sums)
total_relabundance_Genus = rowSums(relabundance_matrix_Genus)

# Get the top highly abundant genera based on relative abundance 
top_Genus_numbers_basic = sort(total_relabundance_Genus, decreasing = TRUE)

# Make into dataframe
top_Genus_numbers_df = as.data.frame(top_Genus_numbers_basic)

# Make into tibble
top_Genus_numbers = as.tibble(top_Genus_numbers_df)

# Rename genera to remove any higher taxonomic names (a quirk of the metaphaln style)
rownames(top_Genus_numbers_df) = gsub(".*(g__*)", "\\1", rownames(top_Genus_numbers_df))

# Get percenatge by dividing by total number of samples (21) and * by 100
top_Genus_pc = top_Genus_numbers %>% 
  mutate(top_Genus_percentage = (top_Genus_numbers_basic/length(colnames(tse_metaphlan_sal_swb))) * 100) %>% 
  mutate(top_Genus = rownames(top_Genus_numbers_df))

# Select only the top 20 genera by relative abundance 
top_Genus_pc$top_20_Genus = ifelse(top_Genus_pc$top_Genus %in% top_Genus, top_Genus_pc$top_Genus, "-other")

# Select only the top genera with a relative abundance above 1%
top_Genus_pc$top_Genus_above_0.1 = ifelse(top_Genus_pc$top_Genus_percentage > 0.1, top_Genus_pc$top_Genus, "-other")
top_Genus_above_0.1 = unique(top_Genus_pc$top_Genus_above_0.1)

# Check significant genera 
sig_res_stage$genus
sig_res_stage$genera_above_0.1pc_relab = ifelse(sig_res_stage$genus %in% top_Genus_above_0.1, sig_res_stage$genus, "-other")
sig_res_stage$genera_above_0.1pc_relab 

# Check which genera are both signifiantly different and highly abundant 
unique(sig_res_stage$genera_above_0.1pc_relab)

# Order the results  and look for highly abundant ones 
sig_res_stage = significant_stage[order(significant_stage$padj),]
sig_res_stage$genus = rownames(sig_res_stage)
head(sig_res_stage)
head(sig_res_stage, n= 15)

# Order the results for stages 1 to 2 
sig_res_stage_1_2 = significant_stage_2_3[order(significant_stage_1_2$padj),]
sig_res_stage_1_2$genus = rownames(sig_res_stage_1_2)
head(sig_res_stage_1_2)
head(sig_res_stage_1_2, n= 15)
nrow(sig_res_stage_1_2)

# Order the results for stages 1 to 2 and find out which are highly abundant (above 0.1%)
sig_res_stage_1_2$genus
sig_res_stage_1_2$genera_above_0.1pc_relab = ifelse(sig_res_stage_1_2$genus %in% top_Genus_above_0.1, sig_res_stage_1_2$genus, "-other")
unique(sig_res_stage_1_2$genera_above_0.1pc_relab)

# Order the results for stages 2 to 3
sig_res_stage_2_3 = significant_stage_2_3[order(significant_stage_2_3$padj),]
sig_res_stage_2_3$genus = rownames(sig_res_stage_2_3)
head(sig_res_stage_2_3)
head(sig_res_stage_2_3, n= 15)
nrow(sig_res_stage_2_3)

# Order the results for stages 2 to 3 and find out which are highly abundant (above 0.1%)
sig_res_stage_2_3$genus
sig_res_stage_2_3$genera_above_0.1pc_relab = ifelse(sig_res_stage_2_3$genus %in% top_Genus_above_0.1, sig_res_stage_2_3$genus, "-other")
unique(sig_res_stage_2_3$genera_above_0.1pc_relab)

# Order the results for stages 3 to 4 
sig_res_stage_3_4 = significant_stage_2_3[order(significant_stage_3_4$padj),]
sig_res_stage_3_4$genus = rownames(sig_res_stage_3_4)
head(sig_res_stage_3_4)
head(sig_res_stage_3_4, n= 15)
nrow(sig_res_stage_3_4)

# Order the results for stages 3 to 4 and find out which are highly abundant (above 0.1%)
sig_res_stage_3_4$genus
sig_res_stage_3_4$genera_above_0.1pc_relab = ifelse(sig_res_stage_3_4$genus %in% top_Genus_above_0.1, sig_res_stage_3_4$genus, "-other")
sig_res_stage_3_4$genera_above_0.1pc_relab
unique(sig_res_stage_3_4$genera_above_0.1pc_relab)

```

## 5.6 Plot counts of genera across noma stages 

First we create a function 

```{r}
plotCountsGGanysig = function(dds, gene, intgroup = "condition", normalized = TRUE,  
                         transform = TRUE, main, xlab = "group", returnData = FALSE,  
                         replaced = FALSE, pc, plot = "point", text = TRUE, 
                         showSignificance = TRUE, sig_method = "deseq", ...) { # <-- ADDED sig_method ARGUMENT
  # Check input gene validity
  stopifnot(length(gene) == 1 & (is.character(gene) | (is.numeric(gene) &  
                                                         (gene >= 1 & gene <= nrow(dds)))))  
  
  # Check if all intgroup columns exist in colData
  if (!all(intgroup %in% names(colData(dds))))  
    stop("all variables in 'intgroup' must be columns of colData")  
  
  # If not returning data, ensure intgroup variables are factors
  if (!returnData) {  
    if (!all(sapply(intgroup, function(v) is(colData(dds)[[v]], "factor")))) {  
      stop("all variables in 'intgroup' should be factors, or choose returnData=TRUE and plot manually")  
    }  
  }  
  
  # Set pseudo count if not provided
  if (missing(pc)) {  
    pc = if (transform) 0.5 else 0  
  }  
  
  # Estimate size factors if missing
  if (is.null(sizeFactors(dds)) & is.null(normalizationFactors(dds))) {  
    dds = estimateSizeFactors(dds)  
  }  
  
  # Get the counts for the gene
  cnts = counts(dds, normalized = normalized, replaced = replaced)[gene, ]  
  
  # Generate grouping variable
  group = colData(dds)[[intgroup]]
  
  # Create the data frame with counts, group, and sample names
  data = data.frame(count = cnts + pc, group = group, sample = colnames(dds), condition = group)  
  
  # Set the plot title
  if (missing(main)) {  
    main = if (is.numeric(gene)) rownames(dds)[gene] else gene
  }  
  
  # Set the y-axis label based on normalization
  ylab = ifelse(normalized, "normalized count", "count")  
  
  # Return the data if requested
  if (returnData) return(data.frame(count = data$count, colData(dds)[intgroup]))  
  
  # Create the base ggplot object
  p = ggplot(data, aes(x = group, y = count, color = group)) +
    labs(x = xlab, y = ylab, title = main) +
    theme_minimal() +
    scale_y_continuous(trans = ifelse(transform, "log10", "identity"))
  
  # Select the type of plot based on the 'plot' argument
  if (plot == "violin") {
    p = p + geom_violin(trim = FALSE) + geom_jitter(shape = 16, position = position_jitter(0.2))
  } else if (plot == "box") {
    p = p + geom_boxplot() + geom_jitter(shape = 16, position = position_jitter(0.2))
  } else {
    p = p + geom_jitter(shape = 16, position = position_jitter(0.2))
  }
  if (text) p = p + geom_text(aes(label = sample), hjust = -0.2, vjust = 0)
  
  # --- MODIFIED SIGNIFICANCE BLOCK ---
  if (showSignificance) {
    # Option 1: Kruskal-Wallis test for comparing groups
    if (sig_method == "kruskal") {
      kw_test = kruskal.test(count ~ group, data = data)
      kw_pval = kw_test$p.value
      p = p + annotate("text", 
                       x = Inf, y = Inf, # Position at top-right
                       hjust = 1.1, vjust = 1.5, # Adjust positioning
                       label = paste("Kruskal-Wallis, p =", round(kw_pval, 2)),
                       size = 4)
    } 
    # Option 2: Original DESeq2 test (only works for 2 groups)
    else if (sig_method == "deseq") {
      if (length(levels(group)) == 2) {
        res = results(dds, contrast = c(intgroup, levels(group)[1], levels(group)[2]), alpha = 0.05)
        res_gene = res[gene, ]
        if (!is.na(res_gene$padj) && res_gene$padj < 0.05) {
          p = p + annotate("text", x = 1.5, y = max(data$count), label = "*", size = 8)
        }
      } else {
        warning("DESeq2 significance test only works for 2 groups. P-value not shown.")
      }
    }
  }
  
  print(p)
}
```

Next we use the function to plot any highly abundant significantly different across noma stages

```{r, warning=FALSE}

Trep_stage = plotCountsGGanysig(dds_stage, gene= "g__Treponema", intgroup = "noma_stage_on_admission", plot = "violin", text = FALSE, showSignificance = TRUE, sig_method = "kruskal")

ggsave("../imgs/Supplementary_Figure_1.png", plot =  Trep_stage, width = 28, height = 16, dpi = 400)

```








